# Preprocessing and filtering

- **filtering strategy**:
    - Filtering is done on size adjusted conts.  
    - We filter on prevalence (keeping taxa with prevalence > `r prevalence`) and taxa_sums() (keeping taxa whose taxa sums/total counts over all samples are above the `r taxa_sums_quantile` quantile)
        - prevalence filter: remove taxa that are only present in very few samples
        - taxa_sums_quantile filter: keep very abundant taxa even if they are only present in a few samples.
    - NG:
        - because of the taxa_sums_quantile filter, the filtering is done on size adjusted counts (Remember: *the size factors that have been calculated above on all taxa before tax_glom*). 
        - this sample size adjustment does not affect prevalence nor relative abundances. So only analyses using ps_tca are affected, not analyses using ps_filt or ps_filt_ra (relative abundance).
    - **NB: Filtering is done on all samples
        

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - filtering  -
# NB: SFs were already calculated in exploratory analysis
ps_tca_filt <- phyloseq::filter_taxa(ps_tca, function(x){
        (sum(x > min_obs) > (prevalence/100)*length(x)) || 
                (sum(x) > quantile(taxa_sums(ps_tca), probs = taxa_sums_quantile/100))
}, prune = TRUE)

ps_filt <- phyloseq::prune_taxa(taxa_names(ps_tca_filt), ps) # you need ps_filt for DESeq2Apply_physeq

ps_filt_ra <- phyloseq::transform_sample_counts(ps_filt, function(x){x/sum(x)})
# --

# - Generate a visual overview of the filtering step -
filterList <- visualize_filtering(physeq = ps_tca, prevalence = prevalence, taxa_sums_quantile = taxa_sums_quantile, phylum_colors = phylum_colors)
# --
```


- taxa within the shaded area were removed by the filtering:

```{r, fig.width = 7, fig.height = 5, warning = FALSE, message = FALSE, echo = FALSE}
filterList[[1]]
# pdf(file = "filter.pdf", width = 7, height = 5)
# filterList[[1]]
# dev.off()
```

```{r, fig.width = 10, fig.height = 8, warning = FALSE, message = FALSE, echo = FALSE}
filterList[[2]]
```


- **`r ntaxa(ps_filt)` out of `r ntaxa(ps)` taxa survived the filter (`r round(100*ntaxa(ps_filt)/ntaxa(ps), 1)` %)**
- **`r sum(otu_table(ps_tca_filt))` out of `r sum(otu_table(ps_tca))` counts survived the filter (`r round(100*sum(otu_table(ps_tca_filt))/sum(otu_table(ps_tca)), 1)` %)** (based on ps_tca vs ps_tca_filt, not ps vs ps_filt)




