# Differential Abundance analysis

- **For all analyses, fisher, DeSeq, wilcoxon: even if no taxon was significantly different between the groups (levels), the ten taxa with the lowest p-value (test stat) are illustrated.**




## Differential prevalence analysis with fisher.test

- **simply a fisher exact test testing whether some taxa are more prevalent in one group than the other**

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps_filt_ra 
# --

# - run the diff_abundance test -
diff_ab_df <- test_diffs_in_prevalence_single(physeq = physeq_to_test, group_var = group_var, compare = group_var_levels, p.adj.method = "fdr", minCount = 0L, symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))
# --

# - catch the hits -
hit_list <- format_hit_table(diff_ab_df, p.adjust.threshold = 0.05, p.adjust.method = NULL)
taxa_hit_df <- hit_list[["hit_table"]]
#--

# - define colors for heatmap -
significance_colors <- brewer.pal(4, "Reds")
significance_colors <- c(rev(significance_colors), "gray", "violet")
names(significance_colors) = c("****", "***", "**", "*", "ns", "?")
taxa_colors <- list("signi_adj" = significance_colors, "Phylum" = phylum_colors)

if (!is.null(shape)){
        sample_colors <- list(color_levels, NA)
        names(sample_colors) <- c(group_var, shape)
} else {
        sample_colors <- list(color_levels)
        names(sample_colors) <- group_var
}
# --

# - get a more informative taxa annotation of the hits -
taxa_annotation <- taxa_hit_df$Annotation
# --

```


- the number of significantly different taxa found: `r hit_list[[2]]`

- list the most differently "abundant" taxa (**NB: even if no taxon is significant the 10 with the lowest p_values are shown. Max 40 are shown**)

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = "asis"}
knitr::kable(head(taxa_hit_df, 40), caption = "taxa with most different prevalence between the groups", row.names = FALSE)
```

- show heat map of the most differently "abundant = prevalent" taxa between the groups

```{r, fig.width = 14, fig.height = 18, warning = FALSE, message = FALSE, echo = FALSE}
p <- plot_heatmap_physeq(physeq_to_test, sample_colors = sample_colors, taxa_info_df = head(taxa_hit_df, 40), taxa_colors = taxa_colors, 
                    taxa_annotation = head(taxa_annotation, 40), max_abundance_for_color = .08, gradient_steps = c(0.15, 0.3, 0.45, 1), 
                    zero_color = "gray", color_function = viridis, color_steps_bw_markers = 10, log_transform = FALSE, drop_color_levels = TRUE,
                    border_color = NA, 
                    cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = TRUE, show_colnames = FALSE, annotation_names_row = FALSE, 
                    annotation_names_col = FALSE, annotation_legend = TRUE, legend = TRUE, font_size = 14, 
                    fontsize_row = 12, fontsize_col = 12, fontsize_number = 12)


# pdf(file = "heatmapprev.pdf", width = 9, height = 6)
# grid::grid.newpage()
# grid::grid.draw(p$gtable)
# dev.off()
```

- **Write your conclusion here:** *e.g: Some taxa were indeed more prevalent in the one group or the other*




## DeSeq2 analysis


```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps_filt 
# --

# - run the test function resulting in result list - NB: a bit special for DESeq because you also get a new physeq
res_list <- test_differential_abundance_DESeq2single(physeq = physeq_to_test, group_var = group_var, compare = group_var_levels, SFs = SFs, p.adjust.method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")), cooksCutoff = TRUE)
diff_ab_df <- res_list[[1]]
physeq_to_test <- res_list[[2]]
# -- 

# - catch the hits -
hit_list <- format_hit_table(diff_ab_df, p.adjust.threshold = 0.05, p.adjust.method = NULL)
taxa_hit_df <- hit_list[["hit_table"]]
#--

# - define colors for heatmap -
significance_colors <- brewer.pal(4, "Reds")
significance_colors <- c(rev(significance_colors), "gray", "violet")
names(significance_colors) = c("****", "***", "**", "*", "ns", "?")
taxa_colors <- list("signi_adj" = significance_colors, "Phylum" = phylum_colors)

if (!is.null(shape)){
        sample_colors <- list(color_levels, NA)
        names(sample_colors) <- c(group_var, shape)
} else {
        sample_colors <- list(color_levels)
        names(sample_colors) <- group_var
}
# --

# - get a more informative taxa annotation of the hits -
taxa_annotation <- taxa_hit_df$Annotation
# --
taxa_annotation1 <- paste(taxa_annotation, " (", taxa_hit_df$signi_adj, ")", sep = "")
# --

# - prepare violin plots -
taxa_illustrated <- 10
violin_plots <- plot_hittaxa_boxAndviolin(physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, taxa_info_df = head(taxa_hit_df, taxa_illustrated), taxa_annotation = head(taxa_annotation1, taxa_illustrated), shape = shape, facet_cols = 5, excludeZero = FALSE, logTransform = FALSE)
# --
```


- the number of significantly different taxa found: `r hit_list[[2]]`

- list the most differently "abundant" taxa (**NB: even if no taxon is significant the 10 with the lowest p_values are shown. Max 40 are shown**)

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = "asis"}
knitr::kable(head(taxa_hit_df, 40), caption = "taxa with most different prevalence between the groups", row.names = FALSE)
```

- show heat map of the most differently abundant taxa between the groups (max 40 are shown)

```{r, fig.width = 14, fig.height = 18, warning = FALSE, message = FALSE, echo = FALSE}
p <- plot_heatmap_physeq(physeq_to_test, sample_colors = sample_colors, taxa_info_df = head(taxa_hit_df, 40), taxa_colors = taxa_colors, 
                    taxa_annotation = head(taxa_annotation, 40), max_abundance_for_color = quantile(as(otu_table(physeq_to_test), "matrix"), 0.993), gradient_steps = c(0.15, 0.3, 0.45, 1), 
                    zero_color = "gray", color_function = viridis, color_steps_bw_markers = 10, log_transform = FALSE, drop_color_levels = TRUE,
                    border_color = NA, 
                    cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = TRUE, show_colnames = FALSE, annotation_names_row = FALSE, 
                    annotation_names_col = FALSE, annotation_legend = TRUE, legend = TRUE, font_size = 14, 
                    fontsize_row = 12, fontsize_col = 12, fontsize_number = 12)

# pdf(file = "heatmapDeseq.pdf", width = 9, height = 6)
# grid::grid.newpage()
# grid::grid.draw(p$gtable)
# dev.off()
```


- show violin plot of the most differently abundant taxa between the groups

```{r, fig.width = 14, fig.height = 8, warning = FALSE, message = FALSE, echo = FALSE}
violin_plots[[2]] + scale_y_log10()

# pdf(file = "heatmap.pdf", width = 7, height = 10)
# grid::grid.newpage()
# grid::grid.draw(p$gtable)
# dev.off()
```


- **Write your conclusion here:** *e.g: Some taxa were indeed more prevalent in the one group or the other*




## Wilcoxon test excluding zeros using relative abundances 

- **NB: done with ps_filt_ra, but could also be done with ps_tca_filt, or logged counts**


```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps_filt_ra
# --

# - run the test function
diff_ab_df <- test_differential_abundance_Wilcoxonsingle(physeq = physeq_to_test, group_var, compare = group_var_levels, excludeZeros = TRUE, p.adjust.method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))

# diff_ab_df1 <- test_differential_abundance_WilcoxonsingleManiCoin(physeq = physeq_to_test, group_var, compare = group_var_levels, block = NULL, excludeZeros = TRUE, p.adjust.method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))

# -- 

# - catch the hits -
hit_list <- format_hit_table(diff_ab_df, p.adjust.threshold = 0.05, p.adjust.method = NULL)
taxa_hit_df <- hit_list[["hit_table"]]
#--


# - define colors for heatmap -
significance_colors <- brewer.pal(4, "Reds")
significance_colors <- c(rev(significance_colors), "gray", "violet")
names(significance_colors) = c("****", "***", "**", "*", "ns", "?")
taxa_colors <- list("signi_adj" = significance_colors, "Phylum" = phylum_colors)

if (!is.null(shape)){
        sample_colors <- list(color_levels, NA)
        names(sample_colors) <- c(group_var, shape)
} else {
        sample_colors <- list(color_levels)
        names(sample_colors) <- group_var
}
# --

# - get a more informative taxa annotation of the hits -
taxa_annotation <- taxa_hit_df$Annotation
# --
taxa_annotation1 <- paste(taxa_annotation, " (", taxa_hit_df$signi_adj, ")", sep = "")
# --

# - prepare violin plots -
taxa_illustrated <- 10
violin_plots <- plot_hittaxa_boxAndviolin(physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, 
                                          taxa_info_df = head(taxa_hit_df, taxa_illustrated), taxa_annotation = head(taxa_annotation1, taxa_illustrated), shape = shape, facet_cols = 5,
                                          excludeZero = TRUE, logTransform = FALSE)
# --
```


- the number of significantly different taxa found: `r hit_list[[2]]`

- list the most differently "abundant" taxa (**NB: even if no taxon is significant the 10 with the lowest p_values are shown. Max 40 are shown**)

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = "asis"}
knitr::kable(head(taxa_hit_df, 40), caption = "taxa with most different prevalence between the groups", row.names = FALSE)
```

- show heat map of the most differently abundant taxa between the groups (max 40 are shown)

```{r, fig.width = 14, fig.height = 18, warning = FALSE, message = FALSE, echo = FALSE}
p <- plot_heatmap_physeq(physeq_to_test, sample_colors = sample_colors, taxa_info_df = head(taxa_hit_df, 40), taxa_colors = taxa_colors, 
                    taxa_annotation = head(taxa_annotation, 40), max_abundance_for_color = quantile(as(otu_table(physeq_to_test), "matrix"), 0.993), gradient_steps = c(0.15, 0.3, 0.45, 1), 
                    zero_color = "gray", color_function = viridis, color_steps_bw_markers = 10, log_transform = FALSE, drop_color_levels = TRUE,
                    border_color = NA, 
                    cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = TRUE, show_colnames = FALSE, annotation_names_row = FALSE, 
                    annotation_names_col = FALSE, annotation_legend = TRUE, legend = TRUE, font_size = 14, 
                    fontsize_row = 12, fontsize_col = 12, fontsize_number = 12)

# pdf(file = "heatmapWilc.pdf", width = 9, height = 6)
# grid::grid.newpage()
# grid::grid.draw(p$gtable)
# dev.off()
```


- show violin plot of the most differently abundant taxa between the groups

```{r, fig.width = 14, fig.height = 8, warning = FALSE, message = FALSE, echo = FALSE}
violin_plots[[2]] + scale_y_log10()

# pdf(file = "violinWilc.pdf", width = 9, height = 6)
# violin_plots[[2]] + scale_y_log10()
# dev.off()
```


- **Write your conclusion here:** *e.g: Some taxa were indeed more prevalent in the one group or the other*


