# Differential Abundance analysis

- **For all analyses, fisher, DeSeq, wilcoxon: even if no taxon was significantly different between the groups (levels), the ten taxa with the lowest p-value (test stat) are illustrated. Also even if more than 25 taxa are significant, only the 25 top hits are illustrated.**
- **NB further: if significance levels are shown with asterisk in the violin plots, these are significant levels coming from a t.test or wilcoxon test (ggpubr stat_compare_mean), the significance levels from the actual test are shown in the tables and indicated in the taxon_names of the violin plots (working on improving this)**



## Differential prevalence analysis with fisher.test

- **simply a fisher exact test testing whether some taxa are more prevalent in one group than the other**

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps_filt_ra 
# --

# - run the diff_abundance test -
diff_ab_df <- test_diffs_in_prevalence_single(physeq = physeq_to_test, group_var = group_var, compare = group_var_levels, p.adj.method = "fdr", minCount = 0L)
# --

# - catch the hits -
hit_list <- format_hit_table(diff_ab_df, p.adjust.threshold = 0.05, p.adjust.method = NULL)
taxa_hit_df <- hit_list[["hit_table"]]
#--

# - define colors for heatmap -
significance_colors <- brewer.pal(4, "Reds")
significance_colors <- c(rev(significance_colors), "gray")
names(significance_colors) = c("****", "***", "**", "*", "ns")
taxa_colors <- list("signi_adj" = significance_colors, "Phylum" = phylum_colors)

if (!is.null(shape)){
        sample_colors <- list(color_levels, NA)
        names(sample_colors) <- c(group_var, shape)
} else {
        sample_colors <- list(color_levels)
        names(sample_colors) <- group_var
}
# --

# - get a more informative taxa annotation of the hits -
taxa_annotation <- get_taxon_names(taxa_hit_df)
# --

# --
```


- the number of significantly different taxa found: `r hit_list[[2]]`

- list the most differently "abundant" taxa (**NB: even if no taxon is significant the 10 with the lowest p_values are shown. Max 40 are shown**)

```{r, warning = FALSE, message = FALSE, echo = FALSE, results = "asis"}
knitr::kable(head(taxa_hit_df, 40), caption = "taxa with most different prevalence between the groups")
```

- show heat map of the most differently "abundant = prevalent" taxa between the groups

```{r, fig.width = 14, fig.height = 18, warning = FALSE, message = FALSE, echo = FALSE}
p <- plot_heatmap_physeq(physeq_to_test, sample_colors = sample_colors, taxa_info_df = head(taxa_hit_df, 40), taxa_colors = taxa_colors, 
                    taxa_annotation = head(taxa_annotation, 40), max_abundance_for_color = .08, gradient_steps = c(0.15, 0.3, 0.45, 1), 
                    zero_color = "gray", color_function = viridis, color_steps_bw_markers = 10, log_transform = FALSE, drop_color_levels = TRUE,
                    border_color = NA, 
                    cluster_cols = FALSE, cluster_rows = FALSE, show_rownames = TRUE, show_colnames = FALSE, annotation_names_row = FALSE, 
                    annotation_names_col = FALSE, annotation_legend = TRUE, legend = TRUE, font_size = 10, 
                    fontsize_row = 6, fontsize_col = 6, fontsize_number = 6)

# pdf(file = "heatmap.pdf", width = 7, height = 10)
# grid::grid.newpage()
# grid::grid.draw(p$gtable)
# dev.off()
```

- **Write your conclusion here:** *e.g: Some taxa were indeed more prevalent in the one group or the other*

