# Analyses at phylum level

- this extra analysis uses the unfiltered phyloseq object. Since these analyses are based on ratio analyses (for example Bacteroidetes to Firmicutes ratio) or relative abundances, they are independent of size factor/library size adjustments


## Phylum relative abundance barplots of the samples

```{r, warning = FALSE, message = FALSE, echo = FALSE}
if (taxonomic_level != "Phylum"){
        ps <- phyloseq::tax_glom(ps, taxrank = "Phylum", NArm = FALSE)
        taxonomic_level <- "Phylum"
}

ps_ra <- phyloseq::transform_sample_counts(ps, function(x){x/sum(x)})


bar_plot_ps_ra <- plot_sample_bars(physeq = ps_ra, x = "Sample", y = "Abundance", group_var = group_var, color_levels = color_levels, fill = "Phylum", color_sample_names = TRUE, col_vec = phylum_colors, facet_grid = NULL)
```

- a plot showing the relative abundance of the phyla

```{r, fig.width = 14, fig.height = 7, warning = FALSE, message = FALSE, echo = FALSE}
bar_plot_ps_ra

# pdf(file = "barRA.pdf", width = 11.5, height = 6)
# bar_plot_ps_ra
# dev.off()
```

- **Write your conclusion here:** *e.g.: often a few samples look very weird because they have almost no Bacteroidetes, or Firmicutes Bacteroidetes difference is clear*




## Firmicutes to other phyla ratios

- *NB again: because the analyses are based on ratios they are independent of library size adjustment and compositionality*

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps
# --

taxa_annotation <- get_taxon_names(as.data.frame(tax_table(ps))[,2:7])
taxa_annotation <- make.unique(taxa_annotation)
taxa_order <- c(names(phylum_colors), taxa_annotation[!taxa_annotation %in% names(phylum_colors)])


# - plot Firmicutes to all other phyla ratio plots NB: you could change taxa_den to maybe only Bacteroidetes -
# NB: if you would like the order based on pValues/significance choose tax_order = NULL
FirmicutesRatioPlots <- plot_taxa_ratios_AllLevels(physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, tax_names = taxa_annotation, taxa_nom = "Firmicutes", taxa_den = NULL, tax_order = taxa_order, test = "wilcox.test", p_adjust_method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))
# NB: you can use the Single version if color_levels is only two levels and you do not want the bars.
# FirmicutesRatioPlots <- plot_taxa_ratiosSingle(physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, tax_names = taxa_annotation, taxa_nom = "Firmicutes", taxa_den = NULL, tax_order = names(phylum_colors), test = "wilcox.test", p_adjust_method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))
# --

```

- **NB: the plot only shows taxa for which the statistical test (t or wilcoxon) was possible, and only samples for which both taxa were present!**
- NB: the plot shows the logged ratios, the analysis was also done on non-logged ratios (wilcoxon significance is the same)


```{r, fig.width = 14, fig.height = 10, warning = FALSE, message = FALSE, echo = FALSE}
FirmicutesRatioPlots[[6]] # in this version I plot the boxplot again the logged ones

# pdf(file = "ratioPlot.pdf", width = 7, height = 7)
# FirmicutesRatioPlots[[6]]
# dev.off()
```

- the p-vals for the ratios

```{r, warning = FALSE, message = FALSE, echo = FALSE}
knitr::kable(FirmicutesRatioPlots[["pValsLog"]]) 
```


- **Write your conclusion here:** *e.g.: The Firmicutes to Proteobacteria seems enriched.*




## Tile plot comparing all phyla against each other

- Basically performs **wilcoxon test** on the abundance ratios of *all phyla combinations*. So similar to the analysis above for Firmicutes ratios, now for all phyla combinations.


```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps
# --

# - calculate ratio matrixes  -
# NB: calculates Matrixes including all samples in physeq (alternative use calculate_raw_TbTmatrixesSingle)
raw_TbTmatrixes <- calculate_raw_TbTmatrixes(physeq = physeq_to_test)
# raw_TbTmatrixes <- lapply(raw_TbTmatrixes, log10)
taxa_annotation <- get_taxon_names(as.data.frame(tax_table(ps))[,2:7])
taxa_annotation <- make.unique(taxa_annotation)
taxa_order <- c(names(phylum_colors), taxa_annotation[!taxa_annotation %in% names(phylum_colors)])
# --


TbT_tile <- create_raw_TbT_TilePlot(TbTmatrixes = raw_TbTmatrixes, physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, signi_level = 0.05, tax_names = taxa_annotation, tax_order = taxa_order, test = "wilcoxon", p_adjust_method = "none")
```


**NB: In the plot: when a phylum combination is colored in the color of a level, the ratio row Phylum/column phylum is higher in that level compared to the other level**


```{r, fig.width = 8, fig.height = 8, warning = FALSE, message = FALSE, echo = FALSE}
TbT_tile

# pdf(file = "tilePlot.pdf", width = 6.2, height = 6)
# TbT_tile
# dev.off()
```

- **Write your conclusion here:** *e.g.: very few ratios differ. Note that the table is symmetrical, it should be.*


