# Analyses at phylum level

- this extra analysis uses the unfiltered phyloseq object

## Phylum relative abundance barplots of the samples

```{r, warning = FALSE, message = FALSE, echo = FALSE}
if (taxonomic_level != "Phylum"){
        ps <- phyloseq::tax_glom(ps, taxrank = "Phylum", NArm = FALSE)
        taxonomic_level <- "Phylum"
}

ps_ra <- phyloseq::transform_sample_counts(ps, function(x){x/sum(x)})


bar_plot_ps_ra <- plot_sample_bars(physeq = ps_ra, x = "Sample", y = "Abundance", group_var = group_var, color_levels = color_levels, fill = "Phylum", color_sample_names = TRUE, col_vec = phylum_colors, facet_grid = NULL)
```

- a plot showing the relative abundance of the phyla

```{r, fig.width = 14, fig.height = 7, warning = FALSE, message = FALSE, echo = FALSE}
bar_plot_ps_ra
```

- **Write your conclusion here:** *e.g.: often a few samples look very weird, or Firmicutes Bacteroidetes difference is clear*



## Firmicutes to other phyla ratios

- *independent of library size adjustment and compositionality*

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps
# --


taxa_annotation <- get_taxon_names(as.data.frame(tax_table(ps)))


# - plot Firmicutes to all other phyla ratio plots NB: you could change taxa_den to maybe only Bacteroidetes -
FirmicutesRatioPlots <- plot_taxa_ratiosSingle(physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, tax_names = taxa_annotation, taxa_nom = "Firmicutes", taxa_den = NULL, tax_order = names(phylum_colors), test = "wilcox.test", p_adjust_method = "fdr", symnum.args = list(cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1), symbols = c("****", "***", "**", "*", "ns")))
# --
```

- NB: the plot only shows taxa for which statistical test (t or wilcoxon) was possible, and only samples for which both taxa were present!
- **NB: the plot shows the logged ratios, the analysis was also done on non-logged ratios, in case of wilcoxon significance is the same**


```{r, fig.width = 14, fig.height = 10, warning = FALSE, message = FALSE, echo = FALSE}
FirmicutesRatioPlots[[6]] # in this version I plot the boxplot again the logged ones
```

- the p-vals for the ratios

```{r, warning = FALSE, message = FALSE, echo = FALSE}
FirmicutesRatioPlots[["pValsLog"]] 
```


- **Write your conclusion here:** *e.g.: The Firmicutes to Proteobacteria seems enriched.*



## Tile plot comparing all phyla against each other

- Basically performs ratio **wilcoxon test** for *all phyla combinations*. 

```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - choose the physeq object to run the test on -
physeq_to_test <- ps
# --

# - calculate ratio matrixes  -
raw_TbTmatrixes <- calculate_raw_TbTmatrixesSingle(physeq = physeq_to_test, group_var = group_var, compare = group_var_levels)
# raw_TbTmatrixes <- lapply(raw_TbTmatrixes, log10)
taxa_annotation <- get_taxon_names(as.data.frame(tax_table(ps)))
# --


TbT_tile <- create_raw_TbT_TilePlotsSingle(raw_TbTmatrixes, physeq = physeq_to_test, group_var = group_var, color_levels = color_levels, signi_level = 0.05, tax_names = taxa_annotation, tax_order = names(phylum_colors), test = "wilcoxon", p_adjust_method = "none")
```


**NB: When colored in a level, the ratio row/column is higher in that level compared to the other level**

```{r, fig.width = 8, fig.height = 8, warning = FALSE, message = FALSE, echo = FALSE}
TbT_tile
```

- **Write your conclusion here:** *e.g.: very few ratios differ. See that the table is symmetrical, it should be.*


