# Alpha diversity analysis 



## raw **alpha-diversity**, i.e. without rarefying


```{r, message = FALSE, warning = FALSE, echo = FALSE}
# - calculate alpha diversity values and add the residuals of a linear fit: alpha-diversity to total_counts, since those fits are often significant -
DF_alpha_list <- calc_alphadiv_plusLmResids(physeq = ps, measures = alpha_div_measures)
lm_fitlist <- DF_alpha_list[[2]]
DF_alpha <- DF_alpha_list[[1]]
# --

# - calculate p-values of t.tests for the different alpha-diversity measures comparing the levels in group_var - 
alpha_div_pVals <- calc_pVals_alphdiv(DF_alpha = DF_alpha, measures = alpha_div_measures, group = group_var, test = "t.test")
# -- 

# - generate boxplots for the different alpha-diversity measures comparing the levels in group_var - 
alpha_div_boxplots <- boxplots_alphdiv(DF_alpha = DF_alpha, measures = alpha_div_measures, group = group_var, shape = shape, color_levels = color_levels, test = "t.test", hide.ns = FALSE)
# --

# - add alpha_diversity_measure vs total counts plots with results of linear fit in title -
alpha_div_lmPlots <- lmPlots_alphdiv(DF_alpha = DF_alpha, lm_fitlist = lm_fitlist, measures = alpha_div_measures, group = group_var, shape = shape, color_levels = color_levels, test = "t.test")
# --

TrList <- c(alpha_div_boxplots, alpha_div_lmPlots)
TrList <- TrList[order(names(TrList))] # to get the plots in the right order
```


```{r, fig.width = 14, fig.height = (length(TrList)/3)*5, warning = FALSE, message = FALSE, echo = FALSE}
do.call("grid.arrange", c(TrList, ncol = 3))
rm(TrList, alpha_div_lmPlots, alpha_div_boxplots)
```

- p.values comparing alpha diversities between the levels in group_var

```{r, warning = FALSE, message = FALSE, echo = FALSE}
knitr::kable(alpha_div_pVals)
# print(xtable(alpha_div_pVals, align = "|c|c|c|c|c|c|c|c|c|", digits = 5), include.rownames = FALSE)
rm(alpha_div_pVals, lm_fitlist, DF_alpha)
```

- **Write your conclusion here:** *e.g.: level 1 has higher alpha diversity. There is a clear trend that alpha diversity gets higher with higher total_counts, but also the residuals of level 1 in a lm alpha diversity to total_counts are higher than those of level 2.*




## alpha diversity after **rarefying**


```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - do the rarefaction using vegan package -
if (is.null(rare_level)) {
        rare_level <- min(sample_sums(ps))
}

count_table_rare <- vegan::rrarefy(as(otu_table(ps), "matrix"), sample = rare_level)

ps.rare <- ps
otu_table(ps.rare) <- otu_table(count_table_rare, taxa_are_rows = taxa_are_rows(ps))
# --
```

- Rarefaction level used: `r rare_level`



```{r, warning = FALSE, message = FALSE, echo = FALSE}
# - calculate alpha diversity values and add the residuals of a linear fit: alpha-diversity to total_counts, since those fits are often highly significant -
DF_alpha_list <- calc_alphadiv_plusLmResids(physeq = ps.rare, measures = alpha_div_measures)
lm_fitlist <- DF_alpha_list[[2]]
DF_alpha <- DF_alpha_list[[1]]
# --

# - calculate p-values of t.tests for the different alpha-diversity measures comparing the levels in group_var - 
alpha_div_pVals <- calc_pVals_alphdiv(DF_alpha = DF_alpha, measures = alpha_div_measures, group = group_var, test = "t.test")
# -- 

# - generate boxplots for the different alpha-diversity measures comparing the levels in group_var - 
alpha_div_boxplots <- boxplots_alphdiv(DF_alpha = DF_alpha, measures = alpha_div_measures, group = group_var, shape = shape, color_levels = color_levels, test = "t.test", hide.ns = FALSE)
# --

# - add alpha_diversity_measure vs total counts plots with results of linear fit in title -
alpha_div_lmPlots <- lmPlots_alphdiv(DF_alpha = DF_alpha, lm_fitlist = lm_fitlist, measures = alpha_div_measures, group = 
                                             group_var, shape = shape, color_levels = color_levels, test = "t.test")
# --

TrList <- c(alpha_div_boxplots, alpha_div_lmPlots)
TrList <- TrList[order(names(TrList))] # to get the plots in the right order

```


```{r, fig.width = 14, fig.height = (length(TrList)/3)*5, warning = FALSE, message = FALSE, echo = FALSE}
do.call("grid.arrange", c(TrList, ncol = 3))
rm(TrList, alpha_div_lmPlots, alpha_div_boxplots)
```

- p.values comparing alpha diversities between the levels in group_var

```{r, warning = FALSE, message = FALSE, echo = FALSE}
knitr::kable(alpha_div_pVals)
# print(xtable(alpha_div_pVals, align = "|c|c|c|c|c|c|c|c|c|", digits = 5), include.rownames = FALSE)
rm(alpha_div_pVals, lm_fitlist, DF_alpha)
```

- **Write your conclusion here:** *e.g.: the linear plots show that after rarefying, all smaples have same sample_sums().*



## Rarefaction curves

- rarefaction curves calculate the alpha diversity (richness) of each sample at different total counts (sample sizes) in steps from 0 up to the chosen rare_max_total value.
    - samples with total counts < rare_max_total have consequently shorter rarefaction curves, i.e. curves that stop before max_total

```{r, warning = FALSE, message = FALSE, echo = FALSE}
if (is.null(rare_max_total)){
        rare_max_total <- quantile(sample_sums(ps), probs = .25)
}

rarefaction_curves <- raref_curve_richness(physeq = ps, group_var = group_var, max_total = rare_max_total, step_size = step_size, color_levels = color_levels, seed = seed)
```



```{r, fig.width = 12, fig.height = 20, warning = FALSE, message = FALSE, echo = FALSE}
grid.arrange(rarefaction_curves[["Tr_richness_col"]], rarefaction_curves[["Tr_richness_group"]], rarefaction_curves[["Tr_richness_grad"]], ncol = 1)
```

- The used rare_max_total was `r rare_max_total`

- **Write your conclusion here:** *e.g.: alpha diversities plateau pretty quickly, most likely not many singletons in.*





